Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-02-02 paper/paper.tex
Remaking the graph for chemo-only surgical manipulations

% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of \citet{simpson_simple_2011} provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in that model is
given instantaneous knowledge of the vector that would move it toward its
pre-specified destination. In addition to the globally supervised
chemoaffinity term, \citet{simpson_simple_2011} introduced a competitive
interaction for space between growth cone agents and a receptor-receptor
axon-axon interaction in order to account for the full set of experimental
manipulations. Here, we propose the replacement of the chemoaffinity term with
a gradient following model consisting of axonal growth cone agents which carry
receptor molecule expression determined by their soma's location of origin on
the retina. Growth cones move on the simulated tectum guided by two pairs of
opposing, orthogonal signalling molecules representing the Ephrin ligands. We
show that with only the receptor-ligand mediated chemoaffinity term and a
receptor-receptor based axon-axon interaction term (meaning that all growth
cone interactions are communicated by receptor signalling), a full range of
experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
an essential requirement for axons to find their
way \citep{gosse_retinotopic_2008} is also accounted for by the model, due to
the opposing influences of signalling gradient pairs. Finally, we demonstrate
that, assuming exponentially varying receptor expression in the retina, tectal
ligand expression should either be exponential if the receptor-ligand signal
induces repulsion (i.e. gradient descent) or logarithmic if the signal induces
attraction (gradient ascent). Thus, we find that a model analogous to the one
we presented in \citet{james_modelling_2020} that accounts for murine barrel
patterning is also a candidate mechanism for the arrangement of the more
continuous retinotectal system.}

@@

The retinotectal projection has been an important model system in the study of
how the cells of the central nervous system are accurately connected together
into functional networks. This projection connects the light-gathering cells
in the retina to movement-related cells in the optic tectum (known as the
superior colliculus in mammals). Light sources originating close to each other
in the environment tend to activate retinal ganglion cells (either
directly \cite{iRGC_citation} or indirectly via rods and cones) situated close
together in the eye so that an image of the environment is formed on the
retinal surface. It has been discovered that the topography of the retina is
preserved within brain regions that process this information such that cells
which are adjacent within the retina primarily excite cells adjacent in the
tectum. This indicates that during development there must exist a mechanism
which ensures the correct arrangement of the axons which leave the retina and
connect to cells in the tectum.

One reason for the success of the study of the retinotectal projection is its
capacity to be experimentally manipulated. In some non-mammalian species, both
the retina and the tectum can be partially ablated, or even physically
reorganised in-vivo, after which axons regrow to restore the order and
function of the system for the individual animal. This manipulability was
exploited in influential work by R. W. Sperry and co-workers during the
mid-twentieth century, leading to Sperry's 1963 summary of
the \emph{chemoaffinity theory} \citep{sperry_chemoaffinity_1963} which
proposes the existence of morphogenetic gradients that guide axons to their
destination. The chemoaffinity theory was given robust support by the
discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995}
which have been shown to form into graded expression fields in the
retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other
sensory systems, such as the somatosensory
system \citep{vanderhaeghen_mapping_2000}. The Ephrin ligands have a clear
effect on axonal outgrowth, as shown in in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and
in-vivo \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.
%
% A-P:
%
% EphA and ephrinAs: Anteroposterior axis of retiontectal projection
% (Nasal-Temporal on retina; Anteroposterior on tectum).
%
% RGM (tectum) Neogenin (retina)
%
% D-V:
%
% EphB and ephrin B: Dorsoventral on retina; dorsoventral on tectum.
%
% En-2: Repels temporal growth cones of xenopus ret-tec axons and attracts
% nasal ones. Brunet et all, Nature, 2005
%
% Ryk/Fz (retina) and Wnt3 (tectum) (Flanahan, 2006)
%

It is tempting to consider the retinotectal projection well understood. With a
comprehensive theory supported by a biochemical mechanism, is there anything
left to understand? That question can be answered by reviewing retinotectal
modelling papers.

Mini-review here, which contrasts some of the modelling papers. The upshot is
that a central problem with models is not how the axons know how to get closer
to their destination, but how they know they have \emph{arrived} at their true
destination. Also make the point that many of the models are phenomenological
in nature and do not elucidate the mechanisms behind the arrangement.

@@

% Visualisation of ligand expression and gradients on the tectum for the
% graft-rotate 90 degrees simulated surgical manipulation
% ./build/sim/agent/tissuevis configs/simpler/m_eE_G.json configs/simpler/e_tecrot90_fig4row1.json
%
\begin{figure}
\includegraphics[width=\linewidth]{./images/Tissuevisb.png}
\caption{Ligand expression and gradients on the tectum for the
graft-rotate 90 degrees simulated surgical manipulation}
\label{f:trot90}
\end{figure}

@@

% ./scripts/fig_GC_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GC_surgical.png}
\caption{$\mathbf{GC}$ model, surgical manipulations}
\label{f:GC}
\end{figure}

@@

@@

% ./scripts/fig_GI_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GI_surgical.png}
\caption{$\mathbf{GI}$ model, surgical manipulations}
\label{f:GI}
\end{figure}

@@

% ./scripts/fig_GJ_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GJ_surgical.png}
\caption{$\mathbf{GJ}$ model, surgical manipulations}
\label{f:GJ}
\end{figure}

@@

>>>>

\color{grey}
% Note this is the SfN draft abstract, and will need rewriting.
\emph{In the classic Chemoaffinity theory, the retinotectal axon projection is
thought to use pairs of orthogonal signalling gradients in the retina to
specify the eventual location of synapses made on the surface of the
tectum/superior colliculus. Similar orthogonal gradients in the tectum provide
a coordinate system which allows the axons to match their prespecified
destination with the correct location. Although the Ephrins have been shown to
guide axons toward their destination, there has yet to emerge a complete
account of the local interactions which halt the axonal growth cones in the
correct locations to recreate the topography of the retinal cells. The model
of \citet{simpson_simple_2011} provides an account of the basic topographic
arrangement of cells on the tectum, as well as reproducing well known surgical
and genetic manipulation experiments. However, it suffers from the absence of
a local chemotactic guidance mechanism. Instead, each agent in that model is
given instantaneous knowledge of the vector that would move it toward its
pre-specified destination. In addition to the globally supervised
chemoaffinity term, \citet{simpson_simple_2011} introduced a competitive
interaction for space between growth cone agents and a receptor-receptor
axon-axon interaction in order to account for the full set of experimental
manipulations. Here, we propose the replacement of the chemoaffinity term with
a gradient following model consisting of axonal growth cone agents which carry
receptor molecule expression determined by their soma's location of origin on
the retina. Growth cones move on the simulated tectum guided by two pairs of
opposing, orthogonal signalling molecules representing the Ephrin ligands. We
show that with only the receptor-ligand mediated chemoaffinity term and a
receptor-receptor based axon-axon interaction term (meaning that all growth
cone interactions are communicated by receptor signalling), a full range of
experimental manipulations to the retinotectal system can be
reproduced. Furthermore, we show that the observation that competition is not
an essential requirement for axons to find their
way \citep{gosse_retinotopic_2008} is also accounted for by the model, due to
the opposing influences of signalling gradient pairs. Finally, we demonstrate
that, assuming exponentially varying receptor expression in the retina, tectal
ligand expression should either be exponential if the receptor-ligand signal
induces repulsion (i.e. gradient descent) or logarithmic if the signal induces
attraction (gradient ascent). Thus, we find that a model analogous to the one
we presented in \citet{james_modelling_2020} that accounts for murine barrel
patterning is also a candidate mechanism for the arrangement of the more
continuous retinotectal system.}
\color{black}

\emph{The development of the axonal projection from the retina to the tectum
is an important model for the study of how the brain organises its
connectivity during development. In fish, amphibians and mammals, retinal
ganglion cells grow their axons into the brain and make topographically
ordered connections on the tectum.
%
Computational models seeking to provide a plausible mechanism by which this
system can order itself have existed since the 1970s. However, most of these
have been phenomenological, some only possible with the invokation of a global
supervising process that provides the state of the entire system to each
axonal agent.
%
Here, we investigated whether a purely self-organising model based on receptor
and ligand interactions could be found, that would successfully reproduce both
the normal development of the projection and its behaviour under the many
experimental manipulations carried out by neurobiologists.
%
We show that a model based on gradient following (via receptor-ligand
interactions) combined with receptor-receptor mediated axonal repulsion can
account for a broad range of experimental results, with information only being
transferred by nearest-neighbour interactions.
}

@@

The retinotectal projection has been an important model system in the study of
how the cells of the central nervous system are accurately connected together
into functional networks. This projection connects the light-gathering cells
in the retina to movement-related cells in the optic tectum (known as the
superior colliculus in mammals). Light sources originating close to each other
in the environment tend to activate retinal ganglion cells (either
directly \cite{iRGC_citation} or indirectly via rods and
cones \cite{rods_cones_citations}) situated close together in the eye so that
an image of the environment is formed on the retinal surface. It has been
discovered that the topography of the retina is preserved within brain regions
that process this information such that cells which are adjacent within the
retina primarily excite cells adjacent in the tectum. This indicates that
during development there must exist a mechanism which ensures the correct
arrangement of the axons which leave the retina and connect to cells in the
tectum.

One reason for the success of the study of the retinotectal projection is its
capacity to be experimentally manipulated. In some non-mammalian species, both
the retina and the tectum can be partially ablated, or even physically
reorganised in-vivo, after which axons regrow to restore the order and
function of the system for the individual animal. Influential work by
R. W. Sperry and co-workers during the mid-twentieth century exploited this
manipulability, leading to Sperry's 1963 summary of the \emph{chemoaffinity
theory} \citep{sperry_chemoaffinity_1963} which proposes the existence of
morphogenetic gradients that guide axons to their destination. The
chemoaffinity theory was given robust support by the discovery of the ephrin
ligands and their
receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been
shown to form into graded expression fields in the
retina \citep{braisted_graded_1997},
tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other
sensory systems, such as the somatosensory
system \citep{vanderhaeghen_mapping_2000}. The Ephrin ligands have a clear
effect on axonal outgrowth, as shown
in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004}
and by
in-vivo \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002}
studies. However, a gradient fo
%
% A-P:
%
% EphA and ephrinAs: Anteroposterior axis of retiontectal projection
% (Nasal-Temporal on retina; Anteroposterior on tectum).
%
% RGM (tectum) Neogenin (retina)
%
% D-V:
%
% EphB and ephrin B: Dorsoventral on retina; dorsoventral on tectum.
%
% En-2: Repels temporal growth cones of xenopus ret-tec axons and attracts
% nasal ones. Brunet et all, Nature, 2005
%
% Ryk/Fz (retina) and Wnt3 (tectum) (Flanahan, 2006)
%

%It is tempting to consider the retinotectal projection well understood. With a
%comprehensive theory supported by a biochemical mechanism, is there anything
%left to understand? That question can be answered by reviewing retinotectal
%modelling papers.

%Mini-review here, which contrasts some of the modelling papers. The upshot is
%that a central problem with models is not how the axons know how to get closer
%to their destination, but how they know they have \emph{arrived} at their true
%destination. Also make the point that many of the models are phenomenological
%in nature and do not elucidate the mechanisms behind the arrangement.%

@@

% Visualisation of ligand expression and gradients on the tectum for the
% graft-rotated 90 degrees simulated surgical manipulation
% ./build/sim/agent/tissuevis configs/simpler/m_eE_G.json configs/simpler/e_tecrot90_fig4row1.json
%
\begin{figure}
\includegraphics[width=\linewidth]{./images/Tissuevisb.png}
\caption{Ligand expression and gradients on the tectum for the
graft-rotate 90 degrees simulated surgical manipulation}
\label{f:trot90}
\end{figure}

@@

% ./scripts/fig_GC_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GC_surgical.png}
\caption{$\mathbf{GC}$ model, surgical manipulations}
\label{f:GCsurg}
\end{figure}

@@

See Fig.\,\ref{f:GI} and Fig.\,\ref{f:GIsurg}.

@@

% ./scripts/fig_GI_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GI_surgical.png}
\caption{$\mathbf{GI}$ model, surgical manipulations}
\label{f:GIsurg}
\end{figure}

@@

See Fig.\,\ref{f:GJ} and Fig.\,\ref{f:GJsurg}.

@@

%
% Supplementary figures in an appendix
%
\appendix

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\section*{Supplementary figures}

% ./scripts/fig_GJ_surgical.sh
\begin{figure}
\includegraphics[width=\linewidth]{./images/fig_GJ_surgical.png}
\caption{$\mathbf{GJ}$ model, surgical manipulations}
\label{f:GJsurg}
\end{figure}
